\paperTitle

### 2014-01-31 doc/icml2014/abstract.tex
Fixed TODO:infinite samples, TODO:using hidden everywhere but latent-variable

Recent work in method of moments provide consistent estimates
  for latent-variable models, avoiding local optima issues,
  but these methods can only be applied to certain types of graphical models.
In this work, we show that the method of moments in conjunction
  with a composite marginal likelihood objective
  yields consistent parameter estimates for a much broader
  class of directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal partial information
about the latent variables, rendering the otherwise non-convex negative
log-likelihood convex.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
Our approach gracefully extends to models outside our class by
  incorporating the partial information via posterior regulraization.
\end{abstract}

>>>>

Recent work in method of moments provide consistent estimates
  for latent-variable models, avoiding local optima issues,
  but these methods can only be applied to certain types of graphical models.
In this work, we show that the method of moments in conjunction
  with a composite marginal likelihood objective
  yields consistent parameter estimates for a much broader
  class of directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal partial information
about the hidden variables, rendering the otherwise non-convex negative
log-likelihood convex.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
Our approach gracefully extends to models outside our class by
  incorporating the partial information via posterior regulraization.
\end{abstract}

### 2014-01-31 doc/icml2014/background.tex
Fixed TODO:infinite samples, TODO:using hidden everywhere but latent-variable

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    \sL &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over latent variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.

@@

To denote the moments, we will use the notation:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k),
\end{align*}
and so on.
We will view $\mO_i, \mO_{ij}, \mO_{ijk}, \cdots$ as tensors of
  order 1, 2, 3, \ldots in $\Re^d, \Re^{d\times d}, \Re^{d \times
  d \times d},\ldots$.
For the latent marginals, we will use 
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a latent variable $h_1$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h) \in \Delta_{k-1}$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

>>>>

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    \sL &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over hidden variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.

@@

To denote the moments, we will use the notation:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k),
\end{align*}
and so on.
We will view $\mO_i, \mO_{ij}, \mO_{ijk}, \cdots$ as tensors of
  order 1, 2, 3, \ldots in $\Re^d, \Re^{d\times d}, \Re^{d \times
  d \times d},\ldots$.
For the hidden marginals, we will use 
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a hidden variable $h_1$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h) \in \Delta_{k-1}$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

### 2014-01-31 doc/icml2014/directed.tex
Fixed TODO:infinite samples, TODO:using hidden everywhere but latent-variable

\todo{need to say we're going to do this for the infinite data case,
and then study bounds}

\todo{try to be consistent: either use latent or hidden}

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  are only a subset of the parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the remaining parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition, we will describe our algorithm entirely in the
  context of directed graphical models, though it generalizes to factor
  graphs.
We also present the algorithm solely in terms of the method of moments; in
  \sectionref{piecewise}, we show how we can get a statistically more
  efficient estimator using convex optimization.

@@

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$, in terms of the latent marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$:
\begin{align*}
\mO_{12} = O \mH_{12} O^\top \quad\Rightarrow\quad
  \mH_{12} = \pinv O \mO_{12} \pinvt O.
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by suitable renormalization.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample complexity}

% PL: "combine" is too vague to be useful
%\LearnMarginals combines two consistent algorithms, \TensorFactorize and
%\LearnCliqueNs, and is thus consistent itself.

>>>>

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  are only a subset of the parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the remaining parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to factor graphs.
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime.
(iii) we present the algorithm solely in terms of the method of moments; in
  \sectionref{piecewise}, we show how we can get a statistically more
  efficient estimator using convex optimization.

@@

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$, in terms of the hidden marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$:
\begin{align*}
\mO_{12} = O \mH_{12} O^\top \quad\Rightarrow\quad
  \mH_{12} = \pinv O \mO_{12} \pinvt O.
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by suitable renormalization.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sample complexity}
\label{sec:sampleComplexity}
% PL: "combine" is too vague to be useful
%\LearnMarginals combines two consistent algorithms, \TensorFactorize and
%\LearnCliqueNs, and is thus consistent itself.

### 2014-01-31 doc/icml2014/intro.tex
Fixed TODO:infinite samples, TODO:using hidden everywhere but latent-variable

% 4. State what we do: exploit moment constraints to make the problem easier.
In this paper,
we show that a much broader class of directed and undirected graphical models can be consistently estimated,
specifically those in which each latent variable has three conditionally
independent observed variables (``views'')---see \propertyref{bottleneck}.
The key idea is to leverage the spectral method of moments,
not to directly provide a consistent parameter estimate as in previous work,
but as partial constraints on a likelihood-based objective.
%For a broad class of directed and undirected graphical models

@@

% 5. We get moments from third-order tensors from bottlenecks and factorize them into marginals,
% then get parameters.
% DONE: try to talk about the approach in terms of three steps
Our approach is illustrated in \figureref{approach}. 
First, we identify the views for each hidden variable $h_i$ (for example,
$x_1^a$, $x_1^b$ and $x_3^a$ are conditionally independent given $h_1$) and use
the tensor factorization algorithm of
\citet{anandkumar12moments,anandkumar13tensor} to identify the \emph{conditional
moments} $\E[x_i^a \mid h_i]$ and $\E[x_i^b \mid h_i]$ for each $i$.
Second, we optimize a \emph{composite marginal likelihood} to recover the marginals over
each latent clique (e.g., $\mathbb P(h_2, h_3, h_4)$).
Normally, such a marginal likelihood objective would be non-convex,
but given the conditional moments, we obtain a convex objective,
which can be globally optimized using EM.
Finally, the latent marginals yield the model parameters
via simple post-processing in the directed case and convex optimization in the undirected case.
%Given these factors, we show that there exists a set of observed
  %variables for every clique such that the likelihood is convex in the
  %parameters of that clique. 
%This guarantees that we can recover the true marginal distributions
  %using EM.
  %finally, with some post-processing, we are able to
  %consistently recover the true parameters of the model.
In \sectionref{directed}, we detail our algorithm and
show that our algorithm yields consistent parameter estimates
under non-degenerate rank assumptions on the conditional moments.
%describe the
  %settings and technical assumptions in which this approach is guaranteed
  %to work.
% PL: hard to talk about $O$ and $T$ at this stage.
%As a particular example, we are able to recover parameters for the
  %model depicted in \figureref{approach} if $O$ and $T$ are full-rank.

%Our approach also applies to undirected log-linear latent variable
  %models, where we use the recovered marginals to make the learning
  %problem convex (\sectionref{undirected}).
Finally, our approach gracefully extends to more general model families
  in which we are unable to recover the marginal distributions for all
  hidden cliques; in such scenarios, we use posterior regularization to constrain
  EM.  
\todo{Is this possible?:  We show that empirically, using such constraints alleviates local
optima issues.}

>>>>

% 4. State what we do: exploit moment constraints to make the problem easier.
In this paper,
we show that a much broader class of directed and undirected graphical models can be consistently estimated,
specifically those in which each hidden variable has three conditionally
independent observed variables (``views'')---see \propertyref{bottleneck}.
The key idea is to leverage the spectral method of moments,
not to directly provide a consistent parameter estimate as in previous work,
but as partial constraints on a likelihood-based objective.
%For a broad class of directed and undirected graphical models

@@

% 5. We get moments from third-order tensors from bottlenecks and factorize them into marginals,
% then get parameters.
% DONE: try to talk about the approach in terms of three steps
Our approach is illustrated in \figureref{approach}. 
First, we identify the views for each hidden variable $h_i$ (for example,
$x_1^a$, $x_1^b$ and $x_3^a$ are conditionally independent given $h_1$) and use
the tensor factorization algorithm of
\citet{anandkumar12moments,anandkumar13tensor} to identify the \emph{conditional
moments} $\E[x_i^a \mid h_i]$ and $\E[x_i^b \mid h_i]$ for each $i$.
Second, we optimize a \emph{composite marginal likelihood} to recover the marginals over
each hidden clique (e.g., $\mathbb P(h_2, h_3, h_4)$).
Normally, such a marginal likelihood objective would be non-convex,
but given the conditional moments, we obtain a convex objective,
which can be globally optimized using EM.
Finally, the hidden marginals yield the model parameters
via simple post-processing in the directed case and convex optimization in the undirected case.
%Given these factors, we show that there exists a set of observed
  %variables for every clique such that the likelihood is convex in the
  %parameters of that clique. 
%This guarantees that we can recover the true marginal distributions
  %using EM.
  %finally, with some post-processing, we are able to
  %consistently recover the true parameters of the model.
In \sectionref{directed}, we detail our algorithm and
show that our algorithm yields consistent parameter estimates
under non-degenerate rank assumptions on the conditional moments.
%describe the
  %settings and technical assumptions in which this approach is guaranteed
  %to work.
% PL: hard to talk about $O$ and $T$ at this stage.
%As a particular example, we are able to recover parameters for the
  %model depicted in \figureref{approach} if $O$ and $T$ are full-rank.

%Our approach also applies to undirected log-linear latent variable
  %models, where we use the recovered marginals to make the learning
  %problem convex (\sectionref{undirected}).
Finally, our approach gracefully extends to more general model families
  in which we are unable to recover the marginal distributions for all
  hidden cliques; in such scenarios, we use posterior regularization to constrain
  EM.  
%\todo{Is this possible?:  We show that empirically, using such constraints alleviates local
%optima issues.}

### 2014-01-31 doc/icml2014/undirected.tex
Fixed TODO:infinite samples, TODO:using hidden everywhere but latent-variable

In this section, we will show that we can extend the results to
  undirected graphical models parameterized as log-linear latent variable
  models.
We exploit a connection between the moments of the clique potentials and
  the log-partition function to show that learning parameters is convex
  given the clique marginals $Z_\sC$. 
The end result is a consistent estimator for log-linear
  latent-variable models which satisfy \propertyref{exclusive-views}.
Finally, we generalize the result to classes when only a subset of the
  clique marginals can be learned.

As before, let $\sG$ be a graphical model with observed variables $X$ and hidden variables $H$.
We assume that the undirected graphical model is a log-linear latent
  variable model with parameters $\theta$,
\begin{align*}
  p_\theta(\vx, \vh) &= \exp( \sum_\sC \theta_\sC^\top \phi_\sC(\vx,\vh) - A(\theta) ),
\end{align*}
where $\phi_\sC(\vx,\vh)$ is a feature vector for the clique $\sC$ and
  $A(\theta) = \sum_{\vx, \vh}  \exp( \sum_\sC \theta_\sC^\top \phi_\sC(\vx,\vh) )$ is the
  log-partition function.
While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
  without loss of generality that $\phi_\sC(x,h)$ are the indicator
  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
  statistics for the model, and hence we do not lose any statistical
  efficiency in using them.}.

>>>>

In this section, we will show that we can extend the results to
  undirected graphical models parameterized as log-linear latent-variable
  models.
We exploit a connection between the moments of the clique potentials and
  the log-partition function to show that learning parameters is convex
  given the clique marginals $Z_\sC$. 
The end result is a consistent estimator for log-linear
  latent-variable models which satisfy \propertyref{exclusive-views}.
Finally, we generalize the result to classes when only a subset of the
  clique marginals can be learned.

As before, let $\sG$ be a graphical model with observed variables $X$ and hidden variables $H$.
We assume that the undirected graphical model is a log-linear latent-variable model with parameters $\theta$,
\begin{align*}
  p_\theta(\vx, \vh) &= \exp( \sum_\sC \theta_\sC^\top \phi_\sC(\vx,\vh) - A(\theta) ),
\end{align*}
where $\phi_\sC(\vx,\vh)$ is a feature vector for the clique $\sC$ and
  $A(\theta) = \sum_{\vx, \vh}  \exp( \sum_\sC \theta_\sC^\top \phi_\sC(\vx,\vh) )$ is the
  log-partition function.
While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
  without loss of generality that $\phi_\sC(x,h)$ are the indicator
  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
  statistics for the model, and hence we do not lose any statistical
  efficiency in using them.}.
